### 2019-01-21 writing/thesis/chapters/methods.tex
Spellcheck and other minor changes.

% TODO: Extract to settings
\chapter{An ADER-DG scheme for the Navier-Stokes Equations}\label{chap:methods}
This chapter describes the numerical and physical background that is needed to simulate reacting fluids with high accuracy.
We start with a description of the arbitrary high order derivatives (\ader{}) discontinuous Galerkin (\dg{}) method, with a focus on equations that contain both advective and diffusive terms.
We then seque into a short description of the \muscl{} scheme, which is a second order finite volume method.

@@

\newcommand{\backgroundPressure}{\overline{\pressure}}
\newcommand{\backgroundRho}{\overline{\Qrho}}
Some of our scenarios can be described as a pertubation of a background state that is initially in hydrostatic balance
\begin{equation}
  \label{eq:hydrostatic-balance}
  \pdv{}{z} \backgroundPressure{\left (z \right )} = -g \backgroundRho(z),
\end{equation}
where $\backgroundPressure$ and $\backgroundRho$ is the background pressure and density.
We now focus on the following part of the momentum equation (neglecting viscous effects) of \cref{eq:equation-set} in combination with gravitational source term~\pcref{eq:source-gravity}
\begin{equation}
  \label{eq:momentum-equation}
  \pdv{\Qj}{t}+ \divergence{ \left(
    \Qv \otimes \Qj + \bm{I} \pressure
    \right)
  }
  =
  -g \bm{k} \cdot \Qrho.
\end{equation}
The background state \cref{eq:hydrostatic-balance} can be comparitively large and can thus diminish other parts of the flux due to numerical instabilities~\cite{muller2010adaptive}.
To avoid this, we split the pressure as $\pressure = \backgroundPressure(z) + \pressure' (\bm{x}, t)$ and the density as $\Qrho = \backgroundRho(z) + \Qrho'(\bm{x}, t)$.
Inserting this splitting and the definition of divergence into \cref{eq:momentum-equation} allows us to apply \cref{eq:hydrostatic-balance}.
The derivatives of the background states are non-zero only for the $z$-direction.
We arrive at 
\begin{equation}
  \label{eq:momentum-equation-split}
  \pdv{\Qj}{t}+ \divergence{ \left(
    \Qv \otimes \Qj + \bm{I} \pressure'
    \right)
  }
  =
  -g \bm{k} \Qrho'.
\end{equation}
We thus cancel parts of the flux with parts of the source term.
This form of the equation is inspired by equation set 3 of~\cite{giraldo2008study}.
They additionally use a splitting of the energy and use the pertubations directly as conservative variables.
%We do not use this to simplify the implementation of gradients.

@@

As a physical boundary condition, we limit ourselves to wall boundary condition.
A standard wall boundary condition for viscous fluid is the no-slip condition, where we assume that when the fluid is close to a wall, it has a velocity of zero relative to the wall.
We enfore this by setting
\begin{align}
  \label{eq:no-slip}
  \begin{split}
  \Qrho^o &= \Qrho^i, \\
  \Qj^o &= 2 \Qj^w - \Qj^i, \\
  \QE^0 &= \QE^i,\\
  \QZ^0 &= \QZ^i,\\
  {(\nabla Q)}^o &= {(\nabla Q^o)},
  \end{split}
\end{align}
where a superscript of $o$ and $i$ denotes the values outside and inside of the domain respectively.
The term $\Qj^w$ denotes the velocity of the wall.

@@

We manually set the energy component of the numerical flux to zero for both no- and free-slip conditions.
\section{The ADER-DG Method}\label{sec:ader-dg}
We describe the arbitrary derivative discontinous Galerkin (\aderdg) method in this chapter.
Our description of the main method follows~\cite{dumbser2008unified,dumbser2010arbitrary,dumbser2018efficient}, our notation follows primarily~\cite{dumbser2018efficient}.
Let $\NVar$ be the number of variables, $\dimensions$ the number of spatial dimensions and $N > 0$ the degree of the basis functions.
We also call $N$ the order of the method.
We discuss the approximation of systems of partial differential equations (\pde) of the form
\begin{equation}
  \label{eq:conservation-law-gradient}
 \frac{\partial}{\partial_t}  \Q + \div \flux(\Q, \gradQ) = \source(\bm{x}, t, \Q),
\end{equation}
which in contrast to hyperbolic conservation laws of the form
\begin{equation}
  \label{eq:conservation-law}
 \frac{\partial}{\partial_t}  \Q + \div \flux(\Q) = \source(\bm{x}, t, \Q)
\end{equation}
contain the gradient $\gradQ \in \mathbb{R}^{\NVar \times \dimensions}$ of the so called vector of conservative variables $\Q \in \mathbb{R}^{\NVar}$~\cite{dumbser2010arbitrary}.
They are therefore not hyperbolic but rather parabolic or elliptic.

We discuss the solution of a conservation law \cref{eq:conservation-law-gradient} with two or three dimensional domain $\domain$ and its boundary $\boundary$.
In our implementation of the discontinous Galerkin (\dg) framework, we approximate this solution in the space
\begin{equation}
  \label{eq:dg-space}
  \broken = \bigcup_k \cell[k]
\end{equation}
of disjoint quadrilateral cells $\cell$.
Note that we do not distinguish between the approximation space and the domain; the use should be clear given its surrounding context.
For each cell we denote its center by $\cellCenter (\cell)$ and its size by $\Delta x, \Delta y$ and $\Delta z$.

@@

The Jacobian determinant of $\mapping$ is simply the volume $V$ of the cell
\begin{equation}
  \label{eq:determinant-mapping}
  \volume = \operatorname{det}(\operatorname{Jacobian}\mapping) = \Delta x \, \Delta y \,\Delta z,
\end{equation}
the Jacobian determinant of $\timeMapping$ is $\Delta t$.
\todo{Similarly -> as well as?}
This is useful for evaluating derivatives by the chain rule and, similarly, integrals by substitution.
For example, we can evaluate a volume integral over a cell by
\begin{equation}
  \label{eq:integration-by-substitution}
  \intdcell{
f(\bm{x})
  }
  =
\volume \intdrefcell{
    \hat{f}(\hat{\bm{x}})
  }.
\end{equation}
\sidetitle{Basis Functions}
The next step is the definition of basis functions for our cells.
We use Langrange polynomials that are defined in the one-dimensional case for the reference interval $(0,1)$ by
\begin{align}
  \label{eq:lagrange-basis}
    \lagrangeRef(x) = \prod_{\substack{0 \leq j \leq N \\ i \neq j}} \frac{x - {\quadNode[j]}}{x_i - \quadNode[j]}.
\end{align}
We can then approximate functions as a sum over the support points $\quadNode[j]$
\begin{equation}
  \label{eq:langrange-expansion-1d}
  f(x) = \sum_{i = 0}^N f(x_i) \lagrangeRef[i](x) = f(x_i) \lagrangeRef[i](x),
\end{equation}
where we used the Einstein summation convention which implies summation over repeated free indices.
This convention is used from here on.
The support points $\quadNode[i]$ are chosen such that they coincide with the nodes of the Gauss-Legendre quadrature of order $N+1$ which we can use to evaluate integrals
\begin{equation}
  \label{eq:quadrature-1d}
  \int_0^1 f(\hat{x}) \dd{\hat{x}} = \sum_{i=0}^{N} \quadWeight[i] f(\quadNode[i]),
\end{equation}
where $\quadWeight[i]$ is the quadrature weight corresponding to the node $\quadNode[i]$.

The basis functions are similarly defined in $\dimensions$-dimensions by considering nodes that are the tensor-products of $d$ one-dimensional Gauss-Legendre quadrature nodes.
They are indexed by a multi-index $n = (n_1, \ldots, n_\dimensions)$
\begin{equation}
  \label{eq:basis-space}
  \sbasisRef[][n] (\bm{x}) = \prod_{i=0}^{\dimensions} \lagrangeRef[n_i] (x_i)
\end{equation}
that is linearised such that $n = (0, \ldots, \sbasisSize)$.
A $d$-dimensional function is then approximated by
\begin{equation}
  \label{eq:lagrange-expansion}
  f(\bm{x}) = \sum_{i = 0}^{\sbasisSize - 1} f(\bm{x}_i) \lagrangeRef[i](\bm{x}) = f(\bm{x}_i) \lagrangeRef[i](\bm{x}).
\end{equation}
Similarly to the one-dimensional case, we can evaluate integrals by
\begin{align}
  \label{eq:quadrature}
  \underbrace{\int_0^1 \cdots \int_0^1}_{\dimensions \text{ times}} f(\hat{\bm{x}}) \dd{\hat{\bm{x}}} &=
 \sum_{n=0}^{\sbasisSize-1} \quadWeight[n] f(\quadNodeNd[n]) \\
  \intertext{with}
  \quadWeight[n] &= \prod_{i}^{d} \quadWeight[n_i],
\end{align}
where the quadrature weights $\quadWeight[n]$ are the product of all relevant one-dimensional quadrature weights, using the aforementioned (linearized) multi-index $n$.

@@

\sidetitle{Expansion of quantities}\todo{Maybe explain this later, right now $\stpredictor$ is not defined!}
All quantities can be expanded in the space-time basis.
For example, the expansion of the space-time predictor $\stpredictor[\cell]$ can be written as
\begin{equation}
  \label{eq:cell-approx-space-time}
  \stpredictor[\cell] (\bm{x}, t) = \stpredictorCoeff[\cell][v',l',n'] \stbasis[\cell][v',l',n'](\bm{x}, t).
\end{equation}
We denote the space-time flux as $\stflux$ and the space-time source as $\stsource$ with coefficient vectors $\stfluxCoeff$ and $\stsourceCoeff$.
Similarly, we define the space degrees of freedoms by an expansion in the space basis, for example for the space predictor
\begin{equation}
  \label{eq:cell-approx-space}
  \spredictor[\cell] (\bm{x}) = \spredictorCoeff[\cell][v',n'] \sbasis[\cell][v',n'] (\bm{x}),
\end{equation}
where the basis does not depend on the time, unlike in \cref{eq:cell-approx-space-time}.
We need the space-expansion for the predictor $\spredictor$, the discrete solution $\ssol$, the source $\ssource$ and the flux $\sflux$,
with coefficients $\spredictorCoeff, \ssolCoeff, \ssourceCoeff$ and $\sfluxCoeff$ respectively.
Each coefficient index generally has the same range as the corresponding basis index.
For example, the indices $n$ and $n'$ both run over the entire space-basis.

@@

Henceforth, we drop the cell index for all quantities and the spatial argument for the basis and test functions.
We integrate the first term of \cref{eq:weak-pde-space-time} by parts in time and the flux divergence in space.
Here we do not use the Riemann solver for the flux boundary term but rather use the known discrete solution at time $t$.
This corresponds to upwinding in time~\cite{dumbser2008unified}.
Note that this neglects the interaction with neighbouring cells.
The cell-local scheme is then
\begin{align}\label{eq:space-time-predictor}
\begin{split}
\intdcell{
  \sttestfunction[][v,l,n] (\nextTime) \stpredictor[i+1](\nextTime)
}
&-
\intdt{\intdcell{
    \pdv{}{t} \sttestfunction[][v,l,n] (t) \stpredictor[i+1](t)
}}
= 
\intdcell{
  \sttestfunction[][v,l,n] (\curTime) \ssol (\curTime)
}
\\&+
\intdt{\intdcell{
    \sttestfunction[][v,l,n] (t) \divergence{\stflux(\stpredictor[i], \gradient{\stpredictor[i]})}
}}
\\&+
\intdt{\intdcell{
    \sttestfunction[][v,l,n] (t) \stsource(\stpredictor[i])
}}.
\end{split}
\end{align}
Here, we introduced the iteration counter $i$ as a subscript for the space-time predictor.
Using the expansions in space and time~\pcref{eq:cell-approx-space,eq:cell-approx-space-time} results in a local systems of equations that can be solved by a fixed-point iteration scheme.
We set the initial value for the space-time predictor as
\begin{equation}
  \label{eq:initial-guess}
\quad \stpredictorCoeff[0][v,l,n] = \ssolCoeff[][v,n]
\end{equation}
which corresponds to a stationary initial guess~\cite{dumbser2008unified}.
For an overview of possible initial guesses, see~\cite{dumbser2018efficient}.

@@

We then integrate the flux divergence by parts in space and insert the Riemann solver to connect with the neighbouring elements.

@@

The timestep is restriced to a so-called \textsc{cfl}-type penalty
\begin{equation}\label{eq:cfl-aderdg}
 %\Delta t \leq  \text{CFL} \, \frac{\alpha(N) \, h}{\maxConvEigen + 2 \maxViscEigen \frac{1}{\alpha(N) h}}
 \Delta t \leq  \text{CFL} \, \alpha(N) \, h \sum_{i=1}^{\dimensions} \frac{1}{\maxConvEigen_i + 2 \maxViscEigen_i \frac{1}{\alpha(N) h}},
\end{equation}
where the eigenvalues are evaluated for the flux Jacobian in direction $i$~\cite{dumbser2010arbitrary,gassner2008discontinuous}.
The constant $\alpha(N) \leq {\left( 2N+1  \right)}^{-1}$ can be obtained from von Neumann analysis on a simple model problem and depends on the approximation order~\cite{dumbser2008unified}.
We use this constant for both hyperbolic and diffusive penalty terms to ensure stability.
For a more detailed and rigorous stability analysis of the Riemann solver, see~\cite{gassner2008discontinuous}.
We use a value of $0.7$ for the constant $\text{CFL}$.

\section{MUSCL-Hancock Finite Volume Scheme}\label{sec:muscl}
We discussed a modern, high-order \dg{} method in the previous chapter.
While this scheme is efficient, it can become unstable, especially in case of discontinuities.
A simple alternative is a finite volume scheme.
We use the \muscl{} method\cite{vanLeer1979towards}.
Our discussion and notation follows the one in~\cite{toro2009riemann}.
Similar to~\cite{toro2009riemann} we only describe the two-dimensional case without source terms, an extension to three dimensions is straigthforward.

In contrast to the previous \dg{} scheme, we now only store averages instead of a full polynomial solution for each cell.
We achieve higher order by reconstructing a linear function from a cell and its neighbours.
Let $\cellAvg$ denote the averages per cell at spatial coordinates $i,j$.

@@

We use the slope to reconstruct the value at the cell boundaries.
In the following, $\cellAvg^{\pm x}$ is the average of the left ($+$) or right ($-$) cell boundary.
Similar, $\cellAvg^{\pm y}$ is the value at the top and bottom cell boundary.
These so called boundary extrapolated values are given by
\newcommand{\extrapolatedCellAvg}[3][i,j]{\cellAvg[#1]^{#3 #2} = \cellAvg #3 \frac{1}{2} \slope{#2}}%
\begin{equation}
\begin{alignedat}{2}
& \extrapolatedCellAvg{x}{-} , \qquad && \extrapolatedCellAvg{x}{+}, \\
& \extrapolatedCellAvg{y}{-} , \qquad && \extrapolatedCellAvg{y}{+}.
\end{alignedat}
\end{equation}
We also use the slopes defined in \cref{eq:slopes} to estimate the gradient of $\cellAvg$ in each cell by the block matrix
\begin{equation}
  \label{eq:muscl-gradient}
  \gradCellAvg = \left( \slope{x} \bigg\rvert \slope{y} \right),
\end{equation}
which is of course an abuse of notation insofar as it is not the gradient of the constant cell value but rather of the linear reconstruction \pcref{eq:slopes}.
\sidetitle{Time Evolution}
To achieve second order in time, we evolve the boundary extrapolated values
\newcommand{\evolvedCellAvg}[2][i,j]{\hat{U}_{#1}^{#2}}
\begin{equation}\label{eq:muscl-time-evolution}
  \begin{split}
    \forall (k \in \{-x, +x, -y, +y \} ): \quad  \hat{U}_{ij}^k &= \cellAvg^k
\\&+ \frac{\Delta t}{2 \Delta x} \left[ \fluxX(\cellAvg^{-x}, \gradCellAvg) - \fluxX(\cellAvg^{+x}, \gradCellAvg) \right]
\\&+ \frac{\Delta t}{2 \Delta y} \left[ \fluxY(\cellAvg^{-y}, \gradCellAvg) - \fluxY(\cellAvg^{+y}, \gradCellAvg) \right],
  \end{split}
\end{equation}
where $\fluxX$ and $\fluxY$ denote the $x$ and $y$ part of the flux respectively.
Note that we do not need to consider neighboring cells for this step.

Finally, the update can be described by
\begin{align}
  \begin{split}
    \cellAvg (t^{n+1}) &= \cellAvg (t^n)
    \\ &+
  \frac{\Delta t}{\Delta x}
    \Riemann \left(
      \evolvedCellAvg[i-1,j]{+x}, \gradCellAvg[i-1,j]; \ %
      \evolvedCellAvg[i,j\phantom{-1}]{-x}, \gradCellAvg[i,j\phantom{-1}]
    \right)
    \\ &-
  \frac{\Delta t}{\Delta x}
    \Riemann \left(
      \evolvedCellAvg[i,j\phantom{+1}]{+x}, \gradCellAvg[i,j\phantom{+1}]; \ %
      \evolvedCellAvg[i+1,j]{-x}, \gradCellAvg[i+1,j]
    \right)
    \\ &+
  \frac{\Delta t}{\Delta y}
    \Riemann \left(
      \evolvedCellAvg[i,j-1]{+y}, \gradCellAvg[i,j-1]; \ %
      \evolvedCellAvg[i,j\phantom{-1}]{-y}, \gradCellAvg[i,j\phantom{-1}]
    \right)
    \\ &-
  \frac{\Delta t}{\Delta y}
    \Riemann \left(
      \evolvedCellAvg[i,j\phantom{+1}]{+y}, \gradCellAvg[i,j\phantom{+1}]; \ %
      \evolvedCellAvg[i,j+1]{-y}, \gradCellAvg[i,j+1]
    \right).
  \end{split}
\end{align}
This results in a scheme that has an order of convergence of two for both space and time.
\todo{Is this enough detail?}
We treat the source term by a simple splitting scheme.
The interested reader is referred to the discussions in~\cite{leVeque2002finite,toro2009riemann}.

@@

Chebychev's inequality

@@

Chebychev's inequality~\pcref{eq:chebychev} then guarantees that only a subset of cells is marked for refinement.

>>>>

% TODO: Extract to settings
\chapter{An ADER-DG scheme for the Navier-Stokes Equations}\label{chap:methods}
This chapter describes the numerical and physical background that is needed to simulate reacting fluids with high accuracy.
We start with a description of the arbitrary high order derivatives (\ader{}) discontinuous Galerkin (\dg{}) method, with a focus on equations that contain both advective and diffusive terms.
We then segue into a short description of the \muscl{} scheme, which is a second order finite volume method.

@@

\newcommand{\backgroundPressure}{\overline{\pressure}}
\newcommand{\backgroundRho}{\overline{\Qrho}}
Some of our scenarios can be described as a perturbation of a background state that is initially in hydrostatic balance
\begin{equation}
  \label{eq:hydrostatic-balance}
  \pdv{}{z} \backgroundPressure{\left (z \right )} = -g \backgroundRho(z),
\end{equation}
where $\backgroundPressure$ and $\backgroundRho$ is the background pressure and density.
We now focus on the following part of the momentum equation (neglecting viscous effects) of \cref{eq:equation-set} in combination with gravitational source term~\pcref{eq:source-gravity}
\begin{equation}
  \label{eq:momentum-equation}
  \pdv{\Qj}{t}+ \divergence{ \left(
    \Qv \otimes \Qj + \bm{I} \pressure
    \right)
  }
  =
  -g \bm{k} \cdot \Qrho.
\end{equation}
The background state \cref{eq:hydrostatic-balance} can be comparatively large and can thus diminish other parts of the flux due to numerical instabilities~\cite{muller2010adaptive}.
To avoid this, we split the pressure as $\pressure = \backgroundPressure(z) + \pressure' (\bm{x}, t)$ and the density as $\Qrho = \backgroundRho(z) + \Qrho'(\bm{x}, t)$.
Inserting this splitting and the definition of divergence into \cref{eq:momentum-equation} allows us to apply \cref{eq:hydrostatic-balance}.
The derivatives of the background states are non-zero only for the $z$-direction.
We arrive at 
\begin{equation}
  \label{eq:momentum-equation-split}
  \pdv{\Qj}{t}+ \divergence{ \left(
    \Qv \otimes \Qj + \bm{I} \pressure'
    \right)
  }
  =
  -g \bm{k} \Qrho'.
\end{equation}
We thus cancel parts of the flux with parts of the source term.
This form of the equation is inspired by equation set 3 of~\cite{giraldo2008study}.
They additionally use a splitting of the energy and use the perturbations directly as conservative variables.
%We do not use this to simplify the implementation of gradients.

@@

As a physical boundary condition, we limit ourselves to wall boundary condition.
A standard wall boundary condition for viscous fluid is the no-slip condition, where we assume that when the fluid is close to a wall, it has a velocity of zero relative to the wall.
We enforce this by setting
\begin{align}
  \label{eq:no-slip}
  \begin{split}
  \Qrho^o &= \Qrho^i, \\
  \Qj^o &= 2 \Qj^w - \Qj^i, \\
  \QE^0 &= \QE^i,\\
  \QZ^0 &= \QZ^i,\\
  {(\nabla Q)}^o &= {(\nabla Q^o)},
  \end{split}
\end{align}
where a superscript of $o$ and $i$ denotes the values outside and inside of the domain respectively.
The term $\Qj^w$ denotes the velocity of the wall.

@@

We manually set the energy component of the numerical flux to zero for both no- and free-slip conditions.
\section{The ADER-DG Method}\label{sec:ader-dg}
We describe the arbitrary derivative discontinuous Galerkin (\aderdg) method in this chapter.
Our description of the main method follows~\cite{dumbser2008unified,dumbser2010arbitrary,dumbser2018efficient}, our notation follows primarily~\cite{dumbser2018efficient}.
Let $\NVar$ be the number of variables, $\dimensions$ the number of spatial dimensions and $N > 0$ the degree of the basis functions.
We also call $N$ the order of the method.
We discuss the approximation of systems of partial differential equations (\pde) of the form
\begin{equation}
  \label{eq:conservation-law-gradient}
 \frac{\partial}{\partial_t}  \Q + \div \flux(\Q, \gradQ) = \source(\bm{x}, t, \Q),
\end{equation}
which in contrast to hyperbolic conservation laws of the form
\begin{equation}
  \label{eq:conservation-law}
 \frac{\partial}{\partial_t}  \Q + \div \flux(\Q) = \source(\bm{x}, t, \Q)
\end{equation}
contain the gradient $\gradQ \in \mathbb{R}^{\NVar \times \dimensions}$ of the so called vector of conservative variables $\Q \in \mathbb{R}^{\NVar}$~\cite{dumbser2010arbitrary}.
They are therefore not hyperbolic but rather parabolic or elliptic.

We discuss the solution of a conservation law \cref{eq:conservation-law-gradient} with two or three dimensional domain $\domain$ and its boundary $\boundary$.
In our implementation of the discontinuous Galerkin (\dg) framework, we approximate this solution in the space
\begin{equation}
  \label{eq:dg-space}
  \broken = \bigcup_k \cell[k]
\end{equation}
of disjoint quadrilateral cells $\cell$.
Note that we do not distinguish between the approximation space and the domain; the use should be clear given its surrounding context.
For each cell we denote its center by $\cellCenter (\cell)$ and its size by $\Delta x, \Delta y$ and $\Delta z$.

@@

The Jacobian determinant of $\mapping$ is simply the volume $V$ of the cell
\begin{equation}
  \label{eq:determinant-mapping}
  \volume = \operatorname{det}(\operatorname{Jacobian}\mapping) = \Delta x \, \Delta y \,\Delta z,
\end{equation}
the Jacobian determinant of $\timeMapping$ is $\Delta t$.
\todo{Similarly -> as well as?}
This is useful for evaluating derivatives by the chain rule and, similarly, integrals by substitution.
For example, we can evaluate a volume integral over a cell by
\begin{equation}
  \label{eq:integration-by-substitution}
  \intdcell{
f(\bm{x})
  }
  =
\volume \intdrefcell{
    \hat{f}(\hat{\bm{x}})
  }.
\end{equation}
\sidetitle{Basis Functions}
The next step is the definition of basis functions for our cells.
We use Lagrange polynomials that are defined in the one-dimensional case for the reference interval $(0,1)$ by
\begin{align}
  \label{eq:lagrange-basis}
    \lagrangeRef(x) = \prod_{\substack{0 \leq j \leq N \\ i \neq j}} \frac{x - {\quadNode[j]}}{x_i - \quadNode[j]}.
\end{align}
We can then approximate functions as a sum over the support points $\quadNode[j]$
\begin{equation}
  \label{eq:langrange-expansion-1d}
  f(x) = \sum_{i = 0}^N f(x_i) \lagrangeRef[i](x) = f(x_i) \lagrangeRef[i](x),
\end{equation}
where we used the Einstein summation convention which implies summation over repeated free indices.
This convention is used from here on.
The support points $\quadNode[i]$ are chosen such that they coincide with the nodes of the Gauss-Legendre quadrature of order $N+1$ which we can use to evaluate integrals
\begin{equation}
  \label{eq:quadrature-1d}
  \int_0^1 f(\hat{x}) \dd{\hat{x}} = \sum_{i=0}^{N} \quadWeight[i] f(\quadNode[i]),
\end{equation}
where $\quadWeight[i]$ is the quadrature weight corresponding to the node $\quadNode[i]$.

The basis functions are similarly defined in $\dimensions$-dimensions by considering nodes that are the tensor-products of $d$ one-dimensional Gauss-Legendre quadrature nodes.
They are indexed by a multi-index $n = (n_1, \ldots, n_\dimensions)$
\begin{equation}
  \label{eq:basis-space}
  \sbasisRef[][n] (\bm{x}) = \prod_{i=0}^{\dimensions} \lagrangeRef[n_i] (x_i)
\end{equation}
that is linearized such that $n = (0, \ldots, \sbasisSize)$.
A $d$-dimensional function is then approximated by
\begin{equation}
  \label{eq:lagrange-expansion}
  f(\bm{x}) = \sum_{i = 0}^{\sbasisSize - 1} f(\bm{x}_i) \lagrangeRef[i](\bm{x}) = f(\bm{x}_i) \lagrangeRef[i](\bm{x}).
\end{equation}
Similarly to the one-dimensional case, we can evaluate integrals by
\begin{align}
  \label{eq:quadrature}
  \underbrace{\int_0^1 \cdots \int_0^1}_{\dimensions \text{ times}} f(\hat{\bm{x}}) \dd{\hat{\bm{x}}} &=
 \sum_{n=0}^{\sbasisSize-1} \quadWeight[n] f(\quadNodeNd[n]) \\
  \intertext{with}
  \quadWeight[n] &= \prod_{i}^{d} \quadWeight[n_i],
\end{align}
where the quadrature weights $\quadWeight[n]$ are the product of all relevant one-dimensional quadrature weights, using the aforementioned (linearized) multi-index $n$.

@@

\sidetitle{Expansion of quantities}
All quantities can be expanded in the space-time basis.
For example, the expansion of the space-time predictor $\stpredictor[\cell]$ can be written as
\begin{equation}
  \label{eq:cell-approx-space-time}
  \stpredictor[\cell] (\bm{x}, t) = \stpredictorCoeff[\cell][v',l',n'] \stbasis[\cell][v',l',n'](\bm{x}, t),
\end{equation}
with coefficients $\stpredictorCoeff$.
The space-time predictor is the result of the predictor step.
We denote the space-time flux as $\stflux$ and the space-time source as $\stsource$ with coefficient vectors $\stfluxCoeff$ and $\stsourceCoeff$.
Similarly, we define the space degrees of freedoms by an expansion in the space basis, for example for the space predictor
\begin{equation}
  \label{eq:cell-approx-space}
  \spredictor[\cell] (\bm{x}) = \spredictorCoeff[\cell][v',n'] \sbasis[\cell][v',n'] (\bm{x}),
\end{equation}
where the basis does not depend on the time, unlike in \cref{eq:cell-approx-space-time}.
We need the space-expansion for the predictor $\spredictor$, the discrete solution $\ssol$, the source $\ssource$ and the flux $\sflux$,
with coefficients $\spredictorCoeff, \ssolCoeff, \ssourceCoeff$ and $\sfluxCoeff$ respectively.
Each coefficient index generally has the same range as the corresponding basis index.
For example, the indices $n$ and $n'$ both run over the entire space-basis.

@@

Henceforth, we drop the cell index for all quantities and the spatial argument for the basis and test functions.
We integrate the first term of \cref{eq:weak-pde-space-time} by parts in time and the flux divergence in space.
Here we do not use the Riemann solver for the flux boundary term but rather use the known discrete solution at time $t$.
This corresponds to upwinding in time~\cite{dumbser2008unified}.
Note that this neglects the interaction with neighboring cells.
The cell-local scheme is then
\begin{align}\label{eq:space-time-predictor}
\begin{split}
\intdcell{
  \sttestfunction[][v,l,n] (\nextTime) \stpredictor[i+1](\nextTime)
}
&-
\intdt{\intdcell{
    \pdv{}{t} \sttestfunction[][v,l,n] (t) \stpredictor[i+1](t)
}}
= 
\intdcell{
  \sttestfunction[][v,l,n] (\curTime) \ssol (\curTime)
}
\\&+
\intdt{\intdcell{
    \sttestfunction[][v,l,n] (t) \divergence{\stflux(\stpredictor[i], \gradient{\stpredictor[i]})}
}}
\\&+
\intdt{\intdcell{
    \sttestfunction[][v,l,n] (t) \stsource(\stpredictor[i])
}}.
\end{split}
\end{align}
Here, we introduced the iteration counter $i$ as a subscript for the space-time predictor.
Using the expansions in space and time~\pcref{eq:cell-approx-space,eq:cell-approx-space-time} results in a local systems of equations that can be solved by a fixed-point iteration scheme.
We set the initial value for the space-time predictor as
\begin{equation}
  \label{eq:initial-guess}
\quad \stpredictorCoeff[0][v,l,n] = \ssolCoeff[][v,n]
\end{equation}
which corresponds to a stationary initial guess~\cite{dumbser2008unified}.
For an overview of possible initial guesses, see~\cite{dumbser2018efficient}.

@@

We then integrate the flux divergence by parts in space and insert the Riemann solver to connect with the neighboring elements.

@@

The timestep is restricted to a so-called \textsc{cfl}-type penalty
\begin{equation}\label{eq:cfl-aderdg}
 %\Delta t \leq  \text{CFL} \, \frac{\alpha(N) \, h}{\maxConvEigen + 2 \maxViscEigen \frac{1}{\alpha(N) h}}
 \Delta t \leq  \text{CFL} \, \alpha(N) \, h \sum_{i=1}^{\dimensions} \frac{1}{\maxConvEigen_i + 2 \maxViscEigen_i \frac{1}{\alpha(N) h}},
\end{equation}
where the eigenvalues are evaluated for the flux Jacobian in direction $i$~\cite{dumbser2010arbitrary,gassner2008discontinuous}.
The constant $\alpha(N) \leq {\left( 2N+1  \right)}^{-1}$ can be obtained from von Neumann analysis on a simple model problem and depends on the approximation order~\cite{dumbser2008unified}.
We use this constant for both hyperbolic and diffusive penalty terms to ensure stability.
For a more detailed and rigorous stability analysis of the Riemann solver, see~\cite{gassner2008discontinuous}.
We use a value of $0.7$ for the constant $\text{CFL}$.

\section{MUSCL-Hancock Finite Volume Scheme}\label{sec:muscl}
We discussed a modern, high-order \dg{} method in the previous chapter.
While this scheme is efficient, it can become unstable, especially in case of discontinuities.
A simple alternative is a finite volume scheme.
We use the \muscl{} method\cite{vanLeer1979towards}.
Our discussion and notation follows the one in~\cite{toro2009riemann}.
Similar to~\cite{toro2009riemann} we only describe the two-dimensional case without source terms, an extension to three dimensions is straightforward.

In contrast to the previous \dg{} scheme, we now only store averages instead of a full polynomial solution for each cell.
We achieve higher order by reconstructing a linear function from a cell and its neighbors.
Let $\cellAvg$ denote the averages per cell at spatial coordinates $i,j$.

@@

We use the slope to reconstruct the value at the cell boundaries.
In the following, $\cellAvg^{\pm x}$ is the average of the left ($+$) or right ($-$) cell boundary.
Similar, $\cellAvg^{\pm y}$ is the value at the top and bottom cell boundary.
These so called boundary extrapolated values are given by
\begin{equation}
\begin{alignedat}{2}
& \extrapolatedCellAvg{x}{-} , \qquad && \extrapolatedCellAvg{x}{+}, \\
& \extrapolatedCellAvg{y}{-} , \qquad && \extrapolatedCellAvg{y}{+}.
\end{alignedat}
\end{equation}
We also use the slopes defined in \cref{eq:slopes} to estimate the gradient of $\cellAvg$ in each cell by the block matrix
\begin{equation}
  \label{eq:muscl-gradient}
  \gradCellAvg = \left( \slope{x} \bigg\rvert \slope{y} \right),
\end{equation}
which is of course an abuse of notation insofar as it is not the gradient of the constant cell value but rather of the linear reconstruction \pcref{eq:slopes}.
\sidetitle{Time Evolution}
To achieve second order in time, we evolve the boundary extrapolated values
\begin{equation}\label{eq:muscl-time-evolution}
  \begin{split}
    \forall (k \in \{-x, +x, -y, +y \} ): \quad  \hat{U}_{ij}^k &= \cellAvg^k
\\&+ \frac{\Delta t}{2 \Delta x} \left[ \fluxX(\cellAvg^{-x}, \gradCellAvg) - \fluxX(\cellAvg^{+x}, \gradCellAvg) \right]
\\&+ \frac{\Delta t}{2 \Delta y} \left[ \fluxY(\cellAvg^{-y}, \gradCellAvg) - \fluxY(\cellAvg^{+y}, \gradCellAvg) \right],
  \end{split}
\end{equation}
where $\fluxX$ and $\fluxY$ denote the $x$ and $y$ part of the flux respectively.
Note that we do not need to consider neighboring cells for this step.

Finally, the update can be described by
\begin{align}
  \begin{split}
    \cellAvg (t^{n+1}) &= \cellAvg (t^n)
    \\ &+
  \frac{\Delta t}{\Delta x}
    \Riemann \left(
      \evolvedCellAvg[i-1,j]{+x}, \gradCellAvg[i-1,j]; \ %
      \evolvedCellAvg[i,j\phantom{-1}]{-x}, \gradCellAvg[i,j\phantom{-1}]
    \right)
    \\ &-
  \frac{\Delta t}{\Delta x}
    \Riemann \left(
      \evolvedCellAvg[i,j\phantom{+1}]{+x}, \gradCellAvg[i,j\phantom{+1}]; \ %
      \evolvedCellAvg[i+1,j]{-x}, \gradCellAvg[i+1,j]
    \right)
    \\ &+
  \frac{\Delta t}{\Delta y}
    \Riemann \left(
      \evolvedCellAvg[i,j-1]{+y}, \gradCellAvg[i,j-1]; \ %
      \evolvedCellAvg[i,j\phantom{-1}]{-y}, \gradCellAvg[i,j\phantom{-1}]
    \right)
    \\ &-
  \frac{\Delta t}{\Delta y}
    \Riemann \left(
      \evolvedCellAvg[i,j\phantom{+1}]{+y}, \gradCellAvg[i,j\phantom{+1}]; \ %
      \evolvedCellAvg[i,j+1]{-y}, \gradCellAvg[i,j+1]
    \right).
  \end{split}
\end{align}
This results in a scheme that has an order of convergence of two for both space and time.
We treat the source term by a simple splitting scheme.
The interested reader is referred to the discussions in~\cite{leVeque2002finite,toro2009riemann}.

@@

Chebyshev's inequality

@@

Chebyshev's inequality~\pcref{eq:chebychev} then guarantees that only a subset of cells is marked for refinement.
